---
title: "Reproducible Research: Peer Assessment 1"
author: "Dana Jacob"
date: "July 11, 2015"
output: html_document
---

library(plyr)

####Read raw data

```{r read_data}
rawData <- read.csv("activity.csv", header = TRUE, sep = ",")
rawData$date <- as.Date(rawData$date)
```

####Calculate total number of steps each day and plot a histogram

```{r total_steps_histogram}
stepsDay <- tapply(rawData$steps, rawData$date, FUN=sum, na.rm=TRUE)
hist(stepsDay, breaks=20, main = "Total number of steps per day",
     xlab = "Number of steps per day", ylab = "Number of days")
```

####Calculate mean and median number of steps per day

```{r mean_median}
meanDay <- mean(stepsDay, na.rm = TRUE)
round(meanDay, digits = 0)
medianDay <- median(stepsDay, na.rm = TRUE)
round(medianDay, digits = 0)
```

Average number of steps per day = 9,354
Median number of steps per day = 10,395

####Plot a time series of average number of steps per 5-minute interval across all days

```{r time_series}
library(ggplot2)

meanInterval <- aggregate(x = list(steps=rawData$steps),
                by = list(interval=rawData$interval), 
                FUN = mean, na.rm = TRUE)

ggplot(meanInterval, aes(x=interval, y=steps)) + 
    geom_line() +
    ggtitle("Average number of steps per interval") +
    xlab("5-minute intervals") +
    ylab("Average number of steps")
```

####Identify the 5-minute interval that has the maximum number of steps

```{r maximum_steps}
maxStep <- meanInterval[which.max(meanInterval$steps),]
round(maxStep, digits = 0)
```
Interval 835 = maximum numberb of steps (206 steps)

####Imputing missing values

Calculate and report the total number of missing values in the dataset 

```{r number_missing}
table (is.na(rawData))
```

There are 2,304 missing values in the data. 

Replace the missing values with the mean for the 5-minute interval and create a new data set

```{r new_data_fill_missing}
newData <- rawData
newData$steps[is.na(newData$steps)] <- 
    tapply(newData$steps, newData$interval, mean, na.rm = TRUE)

#Check to make sure missing values are filled
table(is.na(newData))
```

####Plot a histogram of total number of steps per day using the new dataset

```{r new_total_steps_histogram}
stepsDayNew <- tapply(newData$steps, newData$date, FUN=sum, na.rm=TRUE)
hist(stepsDayNew, breaks=20, main = "Total number of steps per day using new dataset",
     xlab = "Number of steps per day", ylab = "Number of days")
```

The histogram using the new dataset is closer to a mormal distribution

####Calculate mean and median number of steps per day using new dataset

```{r new_mean_median}
meanDayNew <- mean(stepsDayNew, na.rm = TRUE)
round(meanDayNew, digits = 0)
medianDayNew <- median(stepsDayNew, na.rm = TRUE)
round(medianDayNew, digits = 0)
```

Filling in missing values caused the mean and median number of steps per day to increase which makes sense.

####Calculate differences in activity patterns between weekdays and weekends

```{r day_type_factor}
#add a factor variable to the new dataset to indicate whether a given date is a weekday or weekend day

newData$date <- as.Date(newData$date)
newData$dayType <- factor(format(newData$date, "%A"))
levels(newData$dayType) <- list(weekday=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), weekend=c("Saturday", "Sunday"))
```

####Make a panel plot with time series to compare average number of steps per 5-minute interval between weekdays and weekends

```{r panel_plot}
meanIntervalNew <- aggregate(steps ~ interval + dayType,
                             data = newData, FUN = mean)

library(lattice)
xyplot(type = "l", data = meanIntervalNew,
       steps ~ interval | dayType,
       layout = c(1,2), grid = T,
       xlab ="5-minute interval",
       ylab = "Average number of steps")
```

More steps were taken on weekends than on weekdays.





